A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2024-11-11 latex-results/07ClusterEvaluationInterpretation.tex
config updated

In short, we find that the feature-based clustering discerned two
meaningfully different groups of participants. We find an adaptive group
(cluster 1) that reports higher well-being and more positive outgroup
attitudes (\textit{median}) that are also stable over time
(\textit{MAD}, \textit{MAC}) and tend to increase over the 30 day test
period (\textit{linear trend}). This group also reported consistently
more meaningful, need-fulfilling, and cooperative outgroup interactions
(\textit{median}). This group with overwhelmingly positive experiences
stands in contrast with a more detrimental group (cluster 2). This
cluster, on average, reported much less positive, less meaningful, and
less fulfilling interactions and interaction patterns (\textit{median}).
This group also reported less positive outgroup attitudes, lower
well-being, and more discrimination experiences (\textit{median}). At
the same time, for members of this detrimental cluster (cluster 2)
conditions seemed to deteriorate over time (\textit{linear trend}), and
there was generally less consistency in the experiences they were able
to have (\textit{MAC}, \textit{MAD}, \textit{edf}).

To identify these patterns, we first inspect the clusters based on the
average values of meaningful features (see
\fgrref[A]{fig:clusterFeatVar}; \citealp{Kennedy2021}). We see that for
some variables the features are generally stronger in separating the
clusters. We, for example, see that the item on
`\textit{how cooperative the interaction was}' distinguishes the two
clusters across almost all seven features (except for the
\textit{auto-correlation}, see \fgrref[A]{fig:clusterFeatVar}). Compare
this to the `\textit{outgroup attitudes}' item where the differences
between the clusters are much smaller for almost all features. We then
inspect the clusters with a focus on the features (see
\fgrref[B]{fig:clusterFeatVar}). While this is the same data as for the
variable focus, we can see more clearly that some features are better at
distinguishing the clusters across variables. For example, \textit{MAD}
and \textit{median} distinguish the two clusters across almost all
variables (except for the item of whether the interaction was
representative of the outgroup). These two features stand in stark
contrast to other features, such as the
\textit{lag-1 auto correlations}, which showed much smaller differences
between the two clusters (see \fgrref[B]{fig:clusterFeatVar}). This
offers some information on which features were most important in
understanding the two extracted groups. Taking these two perspectives
together, we can also focus on individual features or variables in
particular. We, for example, see a strong difference in the average
well-being, where participants in cluster 2 showed a much lower median
well-being over the time series. At the same time, in terms of
stability, both groups have virtually identical average \textit{MAC}
statistics for well-being (see \fgrref[A]{fig:clusterFeatVar}). There
are, thus, variables and features that distinguish the clusters better
than others and a combination of variables and features lets us explore
meaningful group differences in more detail. In our case, we see that
the central tendency, variability, and linear trend are best at
distinguishing a group with mainly positive experiences (cluster 1) from
a group with a more negative experience (cluster 2). We also see that
our clusters line up with the past literature on the importance of
focusing on simpler and more meaningful statistics
\citep{bringmann2018c, eronen2021a}.

\begin{figure}[!ht] %hbtp
  \caption{Cluster Group Comparisons based on Features and Variables}
  \label{fig:clusterFeatVar}
  \centering\includegraphics[width=\textwidth]{figures/clusterFeatVarComb.pdf}
  \caption*{Note: \\
  "Int." = outgroup interaction. Within the "(B) Feature Focus" subplot, the 'n' and 'Discrimination' comparison variables were not part of the original time series clustering.}
\end{figure}

In the second step, we look at the prototypical trajectories of the
clusters. For k-means clustering it is often recommended to use the
average over time of the responses within the cluster
\citep[see \fgrref{fig:clusterTs};][]{niennattrakul2007}\footnote{It is important to note, however, that direct comparability can be a concern, and often times some subset selection or nonlinear alignment is necessary \citep[e.g.,][]{gupta1996}. Additionally, finding cluster prototypes is often substantially easier with embedded clustering methods because in many cases a cluster-level model is estimated as part of the expectation–maximization procedure \citep[e.g.,][]{denteuling2021} or \textit{S-GIMME} \citep[e.g.][]{lane2019}. For medoid-based clustering algorithms, a common approach is simply using cluster medoid as the prototype \citep{kaufman1990}.}.
Immediately striking are the mean differences, where participants in
cluster 1 had more meaningful and fulfilling outgroup interactions and
also consistently reported more voluntary and cooperative interactions
but fewer accidental and involuntary interactions. The same cluster
(cluster 1) also reported an increase in need-fulfilling interactions
over the 30-day period and an increase in interactions that were
representative of the outgroup. Whereas the other cluster (cluster 2)
showed a decrease in voluntary, cooperative, and positive interactions
over the 30 days. This `deterioration' cluster (cluster 2) also saw a
decrease in general need fulfillment and experienced well-being over the
30 days (see \fgrref[B]{fig:clusterTs}). We also see that while
interaction representativeness, outgroup attitudes, and well-being are
relatively stable for both clusters, the deteriorating cluster (cluster
2) also showed substantially higher variability and instability on most
of the other variables (also see \fgrref[A]{fig:clusterTs}).

Finally, we can also assess the clusters across other individual
difference variables \citep[e.g.,][]{monden2022}. This out-of-feature
comparison allows us to check for data artifacts, as well as check
whether the developmental clusters are associated with important social
markers and individual differences. To illustrate artifact checks, we
added the number of ESM measurements into the comparison and find that
the participants in the deterioration cluster (cluster 2) on average
completed slightly more ESM surveys in general and reported on more
intergroup interactions in particular (see \(n\) in
\fgrref[B]{fig:clusterFeatVar}). In our data exclusion procedures, we
ensured that the general time frame and completion rates are similar for
all participants and indeed the numbers in ESM measurements generally
are largely similar (e.g., see \(n\) for well-being and outgroup
attitudes). However, the difference in the reported number of
interactions might indicate either a clustering artifact or a meaningful
difference. The higher average number of interactions in cluster 2
could, for example, indicate a clustering artifact if variances are
substantially larger due to the larger samples
\citep[e.g., restriction of range in the smaller sample][]{kogan2006}.
In our case, this seems less likely because one out of four variables
did not differ in terms of the MAD (i.e., our selected measurement of
the time series variance; see \fgrref{fig:cluster_comparison_mad} for an
illustration). At the same time, however, the difference in the number
of experienced interactions might also indicate a meaningful difference,
where the deteriorating cluster (cluster 2) on average reported more
outgroup interactions (\textit{difference} = 1.03, \(t\)(150.83) = 7.50,
\(p\) \textless{} .001, \textit{95\%CI} {[}0.76, 1.30{]}), but these
interactions were less voluntary (\textit{difference} = -1.04,
\(t\)(108.89) = -7.71, \(p\) \textless{} .001, \textit{95\%CI} {[}-1.31,
-0.77{]}), less meaningful (\textit{difference} = -1.00, \(t\)(136.40) =
-7.16, \(p\) \textless{} .001, \textit{95\%CI} {[}-1.28, -0.73{]}), and
less positive (\textit{difference} = -1.38, \(t\)(152.31) = -11.94,
\(p\) \textless{} .001, \textit{95\%CI} {[}-1.61, -1.15{]}). Thus, while
more research is needed for a conclusive test, our data seems to suggest
that the differences in reported interactions are a meaningful
difference between the clusters. Such a finding would also be in line
with past research highlighting the role of negative intergroup
interactions in explaining intergroup relations
\citep[e.g.,][]{Barlow2012, Prati2021, Graf2014}.

\begin{figure}[!ht] %hbtp
  \caption{Cluster comparison of Median Absolute Deviation for all variables}
  \label{fig:cluster_comparison_mad}
  \centering\includegraphics[width=\textwidth]{figures/feature_comparison_mad.pdf}
\end{figure}

@@

\begin{figure}[!ht] %hbtp
  \caption{Cluster Group Comparisons over time}
  \label{fig:clusterTs}
  \centering\includegraphics[width=\textwidth]{figures/clusterTsComb.pdf}
  \caption*{Note: \\
  Subplot (A) displays the variable cluster means at every measurement occasion. Subplot (B) shows the GAM spline for each cluster across the measurement occasions. The thinner lines present all individual time series.}
\end{figure}

>>>>

In short, we find that the feature-based clustering discerned two
meaningfully different groups of participants. We find an adaptive group
(cluster 1) that reports higher well-being (\textit{median}:
\textit{difference} = -0.52, \(t\)(153.87) = -3.34, \(p\) = 0.001,
\textit{95\%CI} {[}-0.82, -0.21{]}; also see
\fgrref[A]{fig:cluster_comparison_features}) and more positive outgroup
interactions (\textit{median}: \textit{difference} = -1.38,
\(t\)(152.31) = -11.94, \(p\) \textless{} .001, \textit{95\%CI}
{[}-1.61, -1.15{]}), which are also stable over time (\textit{MAC}:
\textit{difference} = 0.54, \(t\)(153.98) = 3.49, \(p\) \textless{}
.001, \textit{95\%CI} {[}0.23, 0.84{]}) and tend to increase more over
the 30 day test period (\textit{linear trend}: \textit{difference} =
-0.55, \(t\)(149.90) = -3.55, \(p\) \textless{} .001, \textit{95\%CI}
{[}-0.85, -0.24{]}; also see
\fgrref[C]{fig:cluster_comparison_features}). This group also reported
consistently more meaningful (\textit{median}: \textit{difference} =
-1.00, \(t\)(136.40) = -7.16, \(p\) \textless{} .001, \textit{95\%CI}
{[}-1.28, -0.73{]}), need-fulfilling (\textit{median}:
\textit{difference} = -0.99, \(t\)(135.30) = -7.17, \(p\) \textless{}
.001, \textit{95\%CI} {[}-1.26, -0.72{]}), and cooperative outgroup
interactions (\textit{median}: \textit{difference} = -1.33,
\(t\)(120.36) = -11.28, \(p\) \textless{} .001, \textit{95\%CI}
{[}-1.56, -1.10{]}). This group with overwhelmingly positive experiences
stands in contrast with a more detrimental group (cluster 2). This
cluster, on average, reported much less positive, less meaningful, and
less fulfilling interactions and interaction patterns (\textit{median}).
This group also reported less positive outgroup attitudes, lower
well-being, and more discrimination experiences (\textit{median}). At
the same time, for members of this detrimental cluster (cluster 2)
conditions seemed to deteriorate over time (\textit{linear trend}), and
there was generally less consistency in the experiences they were able
to have (\textit{MAC}, \textit{MAD}, \textit{edf}; also see
\fgrref[]{fig:cluster_comparison_features}; for a full and interactive
comparison of all features see \situtorial).

To identify these patterns, we first inspect the clusters based on the
average values of meaningful features (see
\fgrref[A]{fig:clusterFeatVar}; \citealp{Kennedy2021}). We see that for
some variables the features are generally stronger in separating the
clusters. We, for example, see that the item on
`\textit{how cooperative the interaction was}' distinguishes the two
clusters across almost all seven features (except for the
\textit{auto-correlation}, see \fgrref[A]{fig:clusterFeatVar}). Compare
this to the `\textit{outgroup attitudes}' item where the differences
between the clusters are much smaller for almost all features. We then
inspect the clusters with a focus on the features (see
\fgrref[B]{fig:clusterFeatVar}). While this is the same data as for the
variable focus, we can see more clearly that some features are better at
distinguishing the clusters across variables. For example, \textit{MAD}
and \textit{median} distinguish the two clusters across almost all
variables (except for the item of whether the interaction was
representative of the outgroup). These two features stand in stark
contrast to other features, such as the \textit{lag-1 auto correlations}
or the \textit{GAM edf}, which showed much smaller differences between
the two clusters (see \fgrref[B]{fig:clusterFeatVar} and
\fgrref[]{fig:cluster_comparison_features}; please note that we offer
readers an interactive tool to assess the cluster differences for all
features in \situtorial). This offers some information on which features
were most important in differentiating the two extracted groups but also
shows that with real-world data, not all features will have enough range
to distinguish people on all variables (e.g., see the non-linearity
patterns in \fgrref[]{fig:cluster_comparison_features}; for a more
direct illustration of GAM edf differences see \citealp{bringmann2017}).
Taking these two perspectives together, we can also focus on individual
features or variables in particular. We, for example, see a strong
difference in the average well-being, where participants in cluster 2
showed a much lower median well-being over the time series
(\textit{difference} = -0.52, \(t\)(153.87) = -3.34, \(p\) = 0.001,
\textit{95\%CI} {[}-0.82, -0.21{]}). At the same time, in terms of
well-being stability, both groups have virtually identical average
\textit{MAC} statistics for well-being (\textit{difference} = -0.01,
\(t\)(153.96) = -0.04, \(p\) = 0.968, \textit{95\%CI} {[}-0.32, 0.31{]};
also see \fgrref[A]{fig:clusterFeatVar}). There are, thus, variables and
features that distinguish the clusters better than others and a
combination of variables and features lets us explore meaningful group
differences in more detail. In our case, we see that the central
tendency, variability, and linear trend are best at distinguishing a
group with mainly positive experiences (cluster 1) from a group with a
more negative experience (cluster 2). We also see that our clusters line
up with the past literature on the importance of focusing on simpler and
more meaningful statistics
\citep{bringmann2018c, eronen2021a, dejonckheere2019}.

\begin{figure}[!ht] %hbtp
  \caption{Cluster Group Comparisons based on Features and Variables}
  \label{fig:clusterFeatVar}
  \centering\includegraphics[width=\textwidth]{figures/clusterFeatVarComb_tutorial.pdf}
  \caption*{Note: \\
  "Int." = outgroup interaction, "mad" = median absolute deviation, "mac" = mean absolute change, "lin" = linear slope, "edf" = estimated degrees of freedom of an empty GAM spline model, "ar01" = lag-1 autocorrelation, "OCC"/"occ" = out-of-cluster comparison\\
  Within the "(B) Feature Focus" subplot, the 'n (within ooc)' is an out-of-cluster comparison of the within-person available measurements for each variable; the 'between ooc (mean)' are also out-of-cluster comparisons but on a between person level. 'Measurements removed' is the person-specific count of measurement occasions removed during the missingness handling and 'Discrimination' is the scale mean of daily discimination experiences (measured during the final survey).}
\end{figure}

In the second step, we look at the prototypical trajectories of the
clusters. For k-means clustering it is often recommended to use the
average over time of the responses within the cluster
\citep[see \fgrref{fig:clusterTs};][]{niennattrakul2007}\footnote{It is important to note, however, that direct comparability can be a concern, and often times some subset selection or nonlinear alignment is necessary \citep[e.g.,][]{gupta1996}.}.
Immediately striking are the mean differences, where participants in
cluster 1 had more meaningful and fulfilling outgroup interactions and
also consistently reported more voluntary and cooperative interactions
but fewer accidental and involuntary interactions. The same cluster
(cluster 1) also reported an increase in need-fulfilling interactions
over the 30-day period and an increase in interactions that were
representative of the outgroup. Whereas the other cluster (cluster 2)
showed a decrease in voluntary, cooperative, and positive interactions
over the 30 days. This `deterioration' cluster (cluster 2) also saw a
decrease in general need fulfillment but not experienced well-being over
the 30 days (see \fgrref[C]{fig:cluster_comparison_features}). We also
see that while interaction representativeness, outgroup attitudes, and
well-being are relatively stable for both clusters, the deteriorating
cluster (cluster 2) also showed substantially higher variability and
instability on most of the other variables (although these effects are
much smaller; see \fgrref[A]{fig:clusterTs}).

Finally, we can also assess the clusters across other individual
difference variables \citep[e.g.,][]{monden2022}. This out-of-feature
comparison allows us to check for data artifacts, as well as check
whether the developmental clusters are associated with important social
markers and individual differences. To illustrate artifact checks, we
added the number of ESM measurements into the comparison and find that
the participants in the deterioration cluster (cluster 2) on average
completed slightly more ESM surveys in general and reported on more
intergroup interactions in particular (see \(n\) in
\fgrref[B]{fig:clusterFeatVar}). In our data exclusion procedures, we
ensured that the general time frame and completion rates are similar for
all participants and indeed the numbers in ESM measurements generally
are largely similar (e.g., see \(n\) for well-being and outgroup
attitudes). However, the difference in the reported number of
interactions might indicate either a clustering artifact or a meaningful
difference. The higher average number of interactions in cluster 2
could, for example, indicate a clustering artifact if variances are
substantially larger due to the larger samples
\citep[e.g., restriction of range in the smaller sample][]{kogan2006}.
In our case, this seems less likely because one out of four variables
did not differ in terms of the MAD (i.e., our selected measurement of
the time series variance; see \fgrref{fig:cluster_comparison_features}
for an illustration). At the same time, however, the difference in the
number of experienced interactions might also indicate a meaningful
difference, where the deteriorating cluster (cluster 2) on average
reported more outgroup interactions (\textit{difference} = 1.03,
\(t\)(150.83) = 7.50, \(p\) \textless{} .001, \textit{95\%CI} {[}0.76,
1.30{]}), but these interactions were less voluntary
(\textit{difference} = -1.04, \(t\)(108.89) = -7.71, \(p\) \textless{}
.001, \textit{95\%CI} {[}-1.31, -0.77{]}), less meaningful
(\textit{difference} = -1.00, \(t\)(136.40) = -7.16, \(p\) \textless{}
.001, \textit{95\%CI} {[}-1.28, -0.73{]}), and less positive
(\textit{difference} = -1.38, \(t\)(152.31) = -11.94, \(p\) \textless{}
.001, \textit{95\%CI} {[}-1.61, -1.15{]}). Thus, while more research is
needed for a conclusive test, our data seems to suggest that the
differences in reported interactions are a meaningful difference between
the clusters. Such a finding would also be in line with past research
highlighting the role of negative intergroup interactions in explaining
intergroup relations \citep[e.g.,][]{Barlow2012, Prati2021, Graf2014}. A
related validity check was the inclusion of the missingness handling,
where we compared the two clusters on the average number of measurements
removed as part of the missingness handling. We find that the clusters
did not significantly differ on this metric suggesting that the
missingness handling did not affect the cluster separation (also see
\appref[]{app:ValidationAnalyses} and \situtorial).

\begin{figure}[!ht] %hbtp
  \caption{Comparison Cluster Differences by Features and Variables.}
  \label{fig:cluster_comparison_features}
  \centering\includegraphics[width=\textwidth]{figures/feature_comparison_combined.pdf}
  \caption*{Note:\\
  The figure shows the differences between the clusters in the standardized features that were entered into the dimensionality reduction (for each input variable). We display the median (panel A), the median absolute deviation (MAD, panel B), the univariate linear slope (panel C), as well as the estimated degrees of freedom of the generalized additive model splines (GAM edf, panel D). Please also note that as part of \situtorial, we provide readers with an interactive selection tool to compare cluster differences on all variables and features.}
\end{figure}

@@

\begin{figure}[!ht] %hbtp
  \caption{Cluster Group Comparisons over time}
  \label{fig:clusterTs}
  \centering\includegraphics[width=\textwidth]{figures/clusterTsComb.pdf}
  \caption*{Note: \\
  Subplot (A) displays the variable cluster means at every measurement occasion. The thinner lines present all individual time series. Subplot (B) shows the GAM spline for each cluster across the measurement occasions. The thinner lines present all individual GAM Splines.}
\end{figure}

### 2024-11-11 sections/01DataDescription.Rmd
typos fixed
config updated

The original studies included `r sum(missS123$pptFull)` participants ($N_{S1}=$ `r missS123$pptFull[missS123$study==1]`, $N_{S2}=$ `r missS123$pptFull[missS123$study==2]`, $N_{S3}=$ `r missS123$pptFull[missS123$study==3]`) with a total of `r format(sum(missS123$nFull), big.mark=",")` ESM measurements. Each of the studies focused on recently arrived first-generation migrants and each study included a number of idiosyncratic variables relevant for the broader research collective. For our empirical example, we focus on the variables that were collected during the ESM surveys and were available in all three studies. Variable selection and preparation are described as part of the illustration below but for additional methodological details about the study setup see \citet[][]{Kreienkamp2022b}.

>>>>

The original studies included `r sum(missS123$pptFull)` participants ($N_{S1}=$ `r missS123$pptFull[missS123$study==1]`, $N_{S2}=$ `r missS123$pptFull[missS123$study==2]`, $N_{S3}=$ `r missS123$pptFull[missS123$study==3]`) with a total of `r format(sum(missS123$nFull), big.mark=",")` ESM measurements. Each of the studies focused on recently arrived first-generation migrants and each study included a number of idiosyncratic variables relevant for the broader research collective. For our empirical example, we focus on the variables that were collected during ESM surveys and were available in all three studies. Variable selection and preparation are described as part of the illustration below but for additional methodological details about the study setup see \citet[][]{Kreienkamp2022b}. Each study was approved by the ethics board of the university of origin and all participants gave informed consent.

### 2024-11-11 sections/02InputVariables.Rmd
config updated

For our illustration, we include 12 variables that were measured as part of the ESM surveys in all three studies and captured information about the participant's interactions, as well as cognitive-, emotional-, and motivational self in relationship with the majority group (see \tblref{tab:var_selection} for an overview). We chose these aspects in particular because (1) the interaction-specific information exemplified the structural missingness issue of modern ESM data and (2) the motivational, emotional, and cognitive experience offered a diverse conceptualization of migration experience (beyond behavioral measurements) that is becoming more common in the literature \citep[][]{Kreienkamp2022d}. The breadth of the included variables also showcases the utility of the method for a growing body of literature that considers heterogeneous and complex concepts. As a result, the number of included variables is also on the higher end for psychological concepts and additionally allows us to showcase the efficiency benefits of the method and offers a reasonable use case for the optional feature reduction step.

@@

Once the important variables have been selected, the data needs to be prepared for the analysis steps. Importantly, this not only means validating and cleaning the data (e.g., re-coding, removing duplicate or unwanted measurements) but also making the time-series comparable. Two important steps are making the time-frames and response scales comparable across participants --- for example, by choosing a time frame that is common to most participants and standardizing the participants' responses \citep['data exclusion' and 'data transformation' in \fgrref{fig:TSCFlow}; also see][]{liao2005}.

In our illustration data set, the studies differed substantially in the maximum length of participation ($\text{max}(t_{S1})=$ `r missS123$timeFull[missS123$study==1]`, $\text{max}(t_{S2})=$ `r missS123$timeFull[missS123$study==2]`, $\text{max}(t_{S3})=$ `r missS123$timeFull[missS123$study==3]`). This was likely due to the option to continue participation without compensation in the latter study. To make the three studies comparable in participation and time frames, we iteratively removed all measurement occasions and participants that had more than 45% missingness \citep[which was in line with the general recommendation for data that might still need to rely on imputations for later model testing][]{Madley-Dowd2019}. This procedure led to a final sample of `r sum(missS123$pptRed)` participants, who jointly produced `r format(sum(missS123$nRed), big.mark=",")` measurements. Importantly, both the participant response-patterns and the time frame were now substantially more comparable ($\text{max}(t_{S1})=$ `r missS123$timeRed[missS123$study==1]`, $\text{max}(t_{S2})=$ `r missS123$timeRed[missS123$study==2]`, $\text{max}(t_{S3})=$ `r missS123$timeRed[missS123$study==3]`). Full methodological details are available in Online Supplemental Material A, but basic item information, descriptives, and correlations are also available in \tblref{tab:descrLong}.

@@

```{r}
# Sample Descriptives Table --- wider version
allMlCor %>%
  t %>%
  as.data.frame %>%
  kbl(.,
      format = "latex",
      caption = "Correlation Table and Descriptive Statistics",
      label = "descrWide",
      booktabs = TRUE,
      align = c("l", rep("c", ncol(.) - 1))) %>%
  kable_classic() %>%
  add_header_above(
    .,
    c(
      "",
      "Correlations" = ncol(allMlCor),
      "Descriptives" = nrow(allMlCor) - ncol(allMlCor)
    )
  ) %>%
  footnote(
    general = c(
      "Upper triangle: Between-person correlations;",
      "Lower triangle: Within-person correlations;",
      "*** p < .001, ** p < .01,  * p < .05"
    )
  ) %>%
  kable_styling(latex_options = "scale_down") %>%
  gsub("\\begin{table}", "\\begin{sidewaystable}\n\\centering", ., fixed = TRUE) %>%
  gsub("\\end{table}", "\\end{sidewaystable}", ., fixed = TRUE) %>%
  save_kable("tables/descrWide.tex")
```

@@

# Sample Descriptives Table --- longer version
allMlCor %>%
  as.data.frame %>%
  kbl(.,
      format = "latex",
      caption = "Correlation Table and Descriptive Statistics",
      booktabs = TRUE,
      align = "c", #c("l", rep("c", ncol(.) - 1)),
      escape = FALSE,
      label = "descrLong",
      col.names = c(
        #"",
        "\\makecell{Int: \\\\ Accidental}",
        "\\makecell{Int: \\\\ Voluntary}", 
        "\\makecell{Int: \\\\ Cooperative}",
        "\\makecell{Int: \\\\ Representative}",
        "\\makecell{Int: \\\\ Meaningful}",
        "\\makecell{Int: \\\\ Quality}", 
        "\\makecell{Int: \\\\ Need Fulfil.}", 
        "\\makecell{Int: \\\\ Need Fulfil. Partner}", 
        "\\makecell{Attitude \\\\ Partner}",
        "\\makecell{Daytime \\\\ Core Need}",
        "\\makecell{Outgroup \\\\ Attitude}",
        "Well-being"
      )) %>%
  kable_classic() %>%
  pack_rows("Correlations", 1, ncol(allMlCor)) %>%
  pack_rows("Descriptives", ncol(allMlCor)+1, nrow(allMlCor)) %>%
  footnote(
    general = c(
      '"Int." = interaction.',
      'Upper triangle: Between-person correlations;',
      'Lower triangle: Within-person correlations;',
      '*** p < .001, ** p < .01,  * p < .05'
    )
  ) %>%
  kable_styling(latex_options = "scale_down") %>%
  gsub("\\begin{table}", "\\begin{sidewaystable*}[!hbtp]\n\\centering", ., fixed = TRUE) %>%
  gsub("\\end{table}", "\\end{sidewaystable*}", ., fixed = TRUE) %>%
  save_kable("tables/descrLong.tex")
```

>>>>

For our illustration, we include 12 variables that were measured as part of the ESM surveys in all three studies and captured information about the participant's interactions, as well as the cognitive-, emotional-, and motivational self in relationship with the majority group (see \tblref{tab:var_selection} for an overview). We chose these aspects in particular because (1) the interaction-specific information exemplified the structural missingness issue of modern ESM data and (2) the motivational, emotional, and cognitive experience offered a diverse conceptualization of migration experience (beyond behavioral measurements) that is becoming more common in the literature \citep[][]{Kreienkamp2022d}. The breadth of the included variables also showcases the utility of the method for a growing body of literature that considers heterogeneous and complex concepts. As a result, the number of included variables is also on the higher end for psychological concepts and additionally allows us to showcase the efficiency benefits of the method and offers a reasonable use case for the feature reduction step.

@@

descr_N_all <- rbind(
  dtS1Red %>% select(any_of(varNamS123MiRed)) %>% mutate(study = "S1") %>%
    mutate(across(!TID & !study, as.numeric)),
  dtS2Red %>% select(any_of(varNamS123MiRed)) %>% mutate(study = "S2"),
  dtS3Red %>% select(any_of(varNamS123MiRed)) %>% mutate(study = "S3")
)

# Step 1: Count the number of measurements per person within each study
measurement_counts <- descr_N_all %>%
  group_by(study, PID) %>%                # Group by study and PID
  summarise(measurement_count = n()) %>%  # Count the number of rows per PID (this is the number of measurements)
  ungroup()                               # Ungroup the data

# Step 2: Calculate the mean and standard deviation of measurements per study
measurement_summary <- measurement_counts %>%
  group_by(study) %>%                              # Group by study
  summarise(
    min_measurements = min(measurement_count),     # Calculate min of measurements per study
    max_measurements = max(measurement_count),     # Calculate max of measurements per study
    mean_measurements = mean(measurement_count),   # Calculate mean of measurements per study
    sd_measurements = sd(measurement_count)        # Calculate standard deviation of measurements per study
  )

generate_descriptive_report <- function(study_id, measurement_summary) {
  # Extract relevant statistics for the given study
  min_measurements <- measurement_summary$min_measurements[measurement_summary$study == study_id]
  max_measurements <- measurement_summary$max_measurements[measurement_summary$study == study_id]
  mean_measurements <- format(round(measurement_summary$mean_measurements[measurement_summary$study == study_id], 2), nsmall=2)
  sd_measurements <- format(round(measurement_summary$sd_measurements[measurement_summary$study == study_id], 2), nsmall=2)

  # Create the LaTeX formatted string
  latex_string <- paste0(
    "$t_{", study_id, "}$: min = ", min_measurements,
    ", max = ", max_measurements,
    ", mean = ", mean_measurements,
    ", sd = ", sd_measurements
  )

  return(latex_string)
}

latex_S1 <- generate_descriptive_report("S1", measurement_summary)
latex_S2 <- generate_descriptive_report("S2", measurement_summary)
latex_S3 <- generate_descriptive_report("S3", measurement_summary)

Once the important variables have been selected, the data needs to be prepared for the analysis steps. Importantly, this not only means validating and cleaning the data (e.g., re-coding, combining scale items) but also making the time-series comparable. Making the time-frames and response scales comparable across participants, for example, includes choosing a time frame that is common to most participants \citep['data preparation' and 'data cleaning' in \fgrref{fig:TSCFlow}; also see][]{liao2005}.

In our illustration data set, the studies differed substantially in the maximum length of participation ($\text{max}(t_{S1})=$ `r missS123$timeFull[missS123$study==1]`, $\text{max}(t_{S2})=$ `r missS123$timeFull[missS123$study==2]`, $\text{max}(t_{S3})=$ `r missS123$timeFull[missS123$study==3]`). This was likely due to the option to continue participation without compensation in the latter study. To make the three studies comparable in participation and time frames, we iteratively removed all measurement occasions and participants that had more than 45% missingness \citep[which was in line with the general recommendation for data that might still need to rely on imputations for later model testing; see][]{Madley-Dowd2019}\footnote{Please note that for cases where the clustering is the main analysis, this high missingness threshold may be too conservative. As part of our validation analyses in \appref[]{app:ValidationAnalyses} we compare the model presented here with varying levels of missing data allowed.}. This procedure led to a final sample of `r sum(missS123$pptRed)` participants, who jointly produced `r format(sum(missS123$nRed), big.mark=",")` survey responses. Importantly, both the participant response patterns and the time frame were now substantially more comparable (number of measurement occasions per person: `r latex_S1`; `r latex_S2`; `r latex_S3`). It is important to consider that some time series features may be less reliable when the number of measurement occasions per person is low (e.g., below 30 measurements per person), and this should be taken into account when conducting similar analyses. Full methodological details are available in Online Supplemental Material A, but basic item information, descriptives, and correlations are also available in \tblref{tab:descrLong}.

@@

```{r}
# Sample Descriptives Table --- wider version
allMlCor %>%
  t %>%
  as.data.frame %>%
  kbl(.,
      format = "latex",
      caption = "Correlation Table and Descriptive Statistics",
      label = "descrWide",
      booktabs = TRUE,
      align = c("l", rep("c", ncol(.) - 1))) %>%
  kable_classic() %>%
  add_header_above(
    .,
    c(
      "",
      "Correlations" = ncol(allMlCor),
      "Descriptives" = nrow(allMlCor) - ncol(allMlCor)
    )
  ) %>%
  footnote(
    general = c(
      '"Int." = outrgroup interaction, "-GOAL-" = self-identified goal, "-NAME-" = provided name of the interaction partner',
      "Upper triangle: Between-person correlations;",
      "Lower triangle: Within-person correlations;",
      "*** p < .001, ** p < .01,  * p < .05"
    )
  ) %>%
  kable_styling(latex_options = "scale_down") %>%
  gsub("\\begin{table}", "\\begin{sidewaystable}\n\\centering", ., fixed = TRUE) %>%
  gsub("\\end{table}", "\\end{sidewaystable}", ., fixed = TRUE) %>%
  save_kable("tables/descrWide.tex")
```

@@

# Sample Descriptives Table --- longer version
allMlCor %>%
  as.data.frame %>%
  kbl(.,
      format = "latex",
      caption = "Correlation Table and Descriptive Statistics",
      booktabs = TRUE,
      align = "c", #c("l", rep("c", ncol(.) - 1)),
      escape = FALSE,
      label = "descrLong",
      col.names = c(
        #"",
        "\\makecell{Int: \\\\ Accidental}",
        "\\makecell{Int: \\\\ Voluntary}", 
        "\\makecell{Int: \\\\ Cooperative}",
        "\\makecell{Int: \\\\ Representative}",
        "\\makecell{Int: \\\\ Meaningful}",
        "\\makecell{Int: \\\\ Quality}", 
        "\\makecell{Int: \\\\ Need Fulfil.}", 
        "\\makecell{Int: \\\\ Need Fulfil. Partner}", 
        "\\makecell{Attitude \\\\ Partner}",
        "\\makecell{Daytime \\\\ Core Need}",
        "\\makecell{Outgroup \\\\ Attitude}",
        "Well-being"
      )) %>%
  kable_classic() %>%
  pack_rows("Correlations", 1, ncol(allMlCor)) %>%
  pack_rows("Descriptives", ncol(allMlCor)+1, nrow(allMlCor)) %>%
  footnote(
    general = c(
      '"Int." = outgroup interaction, "ICC" = intraclass correlation coefficient, "SD" = standard deviation',
      'Upper triangle: Between-person correlations;',
      'Lower triangle: Within-person correlations;',
      '*** p < .001, ** p < .01,  * p < .05'
    )
  ) %>%
  kable_styling(latex_options = "scale_down") %>%
  gsub("\\begin{table}", "\\begin{sidewaystable*}[!hbtp]\n\\centering", ., fixed = TRUE) %>%
  gsub("\\end{table}", "\\end{sidewaystable*}", ., fixed = TRUE) %>%
  save_kable("tables/descrLong.tex")
```
